# Results {#results}

```{r results-setup, echo=FALSE, message=FALSE, warning=FALSE}
# 1 = green, 2 = blue , 3 = bright blue, 4 = red, 5 = black
ploidy_colours = c("#30a068", "#39bbf7", "#3d85c6", "#cc0000", "#000000") 
real_auto_rate = 0.0056
real_allo_rate = 0.2752

control <- readRDS("_data/control.rds")
simulations <- readRDS("_data/simulations.rds")
source("results.R")

theme_ploidy <- theme(
  legend.position = "top",
  legend.title = element_text(size = 7, face = "bold"),
  legend.text = element_text(size = 7, face = "bold"),
  axis.title = element_text(size = 7, face = "bold"),
  plot.background = element_rect(fill = "#efefef"),
  strip.background = element_rect(colour = "white"),
  strip.text = element_text(size = 7, face = "bold")
)
```

Sterile polyploids were expected to become established at a relative abundance which roughly equalled the rate of nonreduction while this rate was low. However, as the rate of nonreduction was increased up to half the predictability of this relative abundance was expected to vary, either exceeding the rate of nonreduction or decreasing below it. A relative abundance of viable polyploids exceeding half the rate of nonreduction was expected to signal a dissapearance of sterile polyploids and allow fixation of polyploidy to persist. A density dependent effect was also expected to reduce chances of polyploidy fixation for pioneering populations.

The data output by the model was in the form of counts, with each row representing one generation of one simulation. Two datasets are analysed here, for; 1) 100 control simulations, and 2) 1000 experimental simulations. For each dataset, columns recorded the counts of individuals by the following subtypes: seeds, adults, diploids, polyploids, sterile polyploids, and viable polyploids. Sterile and viable polyploids were a subdivision of polyploids, and both diploids and polyploids were a subdivision of adults. Counts recorded were the number of individuals at the end of a generation, after all, transitions were completed. These were used to calculate relative abundance for each subtype. In addition, the experimental variables (rate of nonreduction and starting population size) used for each simulation were recorded in every row, and all simulation instances were separated by identification code.

## General patterns

The pattern of population growth over time when no genome-doubling occurred was consistent and predictable (figure \ref{fig:results-control}). Stability was defined as having a total population that reached carrying capacity (K) and persisted around that point for the entire simulation. Fixation was defined as either diploids or polyploids being the only ploidy type persisting on the landscape. Without nonreduction, whatever the starting population size, the number of adults quickly stabilised in under 15 generations (mean = 3.07, sd = 2.52, N = 100) and persisted in a stable manner for the entire 1000 generations. When nonreduction was included in the experimental simulations, this stability gradually became disrupted as polyploids appeared in the system and began to spread. Broad patterns observed in the results could be split into four categories based on their outcome (figure \ref{fig:results-hist}): 

1. Stable coexistence of a mixed diploid-polyploid population.
2. Complete extinction of both diploids and polyploids.
3. Unstable fixation of polyploidy. 
4. Stable fixation of polyploidy. 

Diploids never became fixated when nonreduction was enabled, as new polyploids arising from diploid progenitors would always keep appearing in the system. This process did not work the other way around; polyploids did not give rise to diploids.

```{r results-control, dev="png", dpi=300, fig.height=2, fig.cap="\\label{fig:results-control}\\textbf{Speed of diploid stabilisation for 100 control simulations:} stability was defined as having a total population that reached carrying capacity (K) and persisted around that point for the entire simulation. Each point represents one simulation. Starting population frequency is the number of adults at the beginning of the simulaiton relative to K, making the scale pioneering (close to zero) to established (close to 1)."}
outcomes %>%
  filter(outcome == "control") %>%
  ggplot2::ggplot(aes(x = starting_N/100000, y = gen_reached)) +
    geom_hline(yintercept = 3.07, linetype = "dotted") +
    geom_point(alpha = .5, size = 1) +
    theme_classic() + theme_ploidy +
    xlab("Starting population frequency") +
    ylab("Generation stability reached") +
    scale_x_continuous(breaks = seq(0, 1, .5)) +
    scale_y_continuous(limits = c(0, 15), breaks = seq(0, 15, 5)) +
    scale_colour_manual(
      values = ploidy_colours[c(5,1,4,3,2)],
      name = "Outcome",
      guide = guide_legend(ncol = 2)
    ) +
    theme(legend.position = "none")
```

```{r results-hist, dev="png", dpi=300, fig.height=3, fig.cap="\\label{fig:results-hist}textbd{Simulations subset by outcome:} Each bin shows the number of simulations (from 1000 instances) within that group of nonreduction rates that had a particular outcome. Simulations ending with stable coexistence (top left) had a total adult population around the carrying capacity (100000) with both diploids and polyploids surviving throughout all 1000 generations. Simulations with total extinction (top right) had neither diploid or adults surviving to 1000 generations. Those where the relative abundance of polyploids persisting through to 1000 generations was 1, were considered fixated. Simulations ending in unstable polyploid fixation (bottom left) had polyploid populations of only 1 or 2 individuals (0.00001-0.00002% of the carrying capacity). Simulations ending with stable polyploid fixation (bottom right) had a total adult population around the carrying capacity (100000) which survived through to 1000 generations. The dotted red line marks the threshold rate of nonreduction (0.152) which divided outcome probability."}
outcomes %>%
  filter(outcome != "control") %>%
  ggplot(aes(x = ploidy_rate)) + 
    geom_histogram(colour = "white", bins = 50) +
    geom_vline(xintercept = 0.152, colour = ploidy_colours[4], linetype = "dotted") +
    scale_fill_manual(values = ploidy_colours[c(1,4,3,2)], name = "") +
    xlab("Rate of nonreduction") +
    ylab("Number of simulations") +
    theme_classic() + theme_ploidy +
    facet_wrap(~outcome, ncol = 2)
```

Below nonreduction at rates of 0.152, sterile polyploids coexisted with diploids and viable polyploids; all became established at varying relative abundances so that the total population stabilised around K. In figure \ref{fig:results-coexistence} this is visualised over time with the broad areas of dark dots that persist consistently across the length of the x-axis, representing all the simulations that achieved stable coexistence. Above nonreduction rates of 0.152, the only simulations in which polyploids were able to persist were those where only viable polyploids remained (figures \ref{fig:results-stable-fixation} and \ref{fig:results-unstable-fixation}). Increasing nonreduction rate caused booming sterile and viable polyploid populations to monopolise the landscape. Sterile polyploid populations peaked first, quickly followed by viable polyploids just a few generations later. Viable polyploids tended to become more abundant than sterile polyploids with relative abundaces over 0.5. The booming rise of polyploidy (in both sterile and viable forms) disrupted the production and survival of diploids in the system, causing population instability that most frequently led to total extinction \ref{fig:results-extinction}. 

```{r results-time-1, dev="png", dpi=300, fig.height=6, fig.cap="\\label{fig:results-coexistence}\\textbf{Coexistence:} changes in relative diploid-polyploid abundances that culminate in coexistence. Each point represents the 1 of 1000 simulations. Points are coloured from black to green based on the rate of nonreduction; simulations that ran with higher rates are brighter green. The dotted black line highlights a frequency of half the population capacity (0.5).", warning=FALSE}
subset_labels <- c("diploids", "sterile polyploids", "viable polyploids")
names(subset_labels) <- c("diploid_adults", "sterile_polyploid_adults", "viable_polyploid_adults")
            
simulations %>%
  filter(ID %in% stability$ID) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::mutate(
    viable_polyploid_adults = polyploid_adults - sterile_polyploid_adults
  ) %>%
  tidyr::gather(subset, count, c(diploid_adults, sterile_polyploid_adults, viable_polyploid_adults)) %>%
  dplyr::filter(count != 0) %>%
  ggplot(aes(x = gen, y = (count*100000)/(adults*100000), colour = ploidy_rate)) +
  geom_hline(yintercept = .5, linetype = "dotted") +
  geom_point(alpha = 1, size = .5) +
  scale_colour_gradient(
    low = "black", high = ploidy_colours[1], 
    name = "Rate of\nnonreduction",
    limits = c(0, 0.5)
  ) +
  theme_classic() + theme_ploidy +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(breaks = seq(0, 1, 0.5)) +
  xlab("Time (generations)") +
  ylab("Relative abundance") +
  facet_wrap(~subset, ncol=1, labeller = labeller(subset = subset_labels))
```

```{r results-time-2, dev="png", dpi=300, fig.height=6, fig.cap="\\label{fig:results-extinction}\\textbf{Extinction:} changes in relative diploid-polyploid abundances that culminate in extinction. Each point represents the 1 of 1000 simulations. Points are coloured from black to green based on the rate of nonreduction; simulations that ran with higher rates are brighter green. The dotted black line highlights a frequency of half the population capacity (0.5).", warning=FALSE}
subset_labels <- c("diploids", "sterile polyploids", "viable polyploids")
names(subset_labels) <- c("diploid_adults", "sterile_polyploid_adults", "viable_polyploid_adults")
            
simulations %>%
  filter(ID %in% extinctions$ID) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::mutate(
    viable_polyploid_adults = polyploid_adults - sterile_polyploid_adults
  ) %>%
  tidyr::gather(subset, count, c(diploid_adults, sterile_polyploid_adults, viable_polyploid_adults)) %>%
  dplyr::filter(count != 0) %>%
  ggplot(aes(x = gen, y = (count*100000)/(adults*100000), colour = ploidy_rate)) +
  geom_hline(yintercept = .5, linetype = "dotted") +
  geom_point(alpha = 1, size = .5) +
  scale_colour_gradient(
    low = "black", high = ploidy_colours[1], 
    name = "Rate of\nnonreduction",
    limits = c(0, 0.5)
  ) +
  theme_classic() + theme_ploidy +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(breaks = seq(0, 1, 0.5)) +
  xlab("Time (generations)") +
  ylab("Relative abundance") +
  facet_wrap(~subset, ncol=1, labeller = labeller(subset = subset_labels))
```

```{r results-time-3, dev="png", dpi=300, fig.height=6, fig.cap="\\label{fig:results-unstable-fixation}\\textbf{Unstable polyploid fixation:} changes in relative diploid-polyploid abundances that culminate in unstable polyploid fixation. Each point represents the 1 of 1000 simulations. Points are coloured from black to green based on the rate of nonreduction; simulations that ran with higher rates are brighter green. The dotted black line highlights a frequency of half the population capacity (0.5).", warning=FALSE}
subset_labels <- c("diploids", "sterile polyploids", "viable polyploids")
names(subset_labels) <- c("diploid_adults", "sterile_polyploid_adults", "viable_polyploid_adults")
            
simulations %>%
  filter(ID %in% unstable_polyploid_winners$ID) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::mutate(
    viable_polyploid_adults = polyploid_adults - sterile_polyploid_adults
  ) %>%
  tidyr::gather(subset, count, c(diploid_adults, sterile_polyploid_adults, viable_polyploid_adults)) %>%
  dplyr::filter(count != 0) %>%
  ggplot(aes(x = gen, y = (count*100000)/(adults*100000), colour = ploidy_rate)) +
  geom_hline(yintercept = .5, linetype = "dotted") +
  geom_point(alpha = 1, size = .5) +
  scale_colour_gradient(
    low = "black", high = ploidy_colours[1], 
    name = "Rate of\nnonreduction",
    limits = c(0, 0.5)
  ) +
  theme_classic() + theme_ploidy +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(breaks = seq(0, 1, 0.5)) +
  xlab("Time (generations)") +
  ylab("Relative abundance") +
  facet_wrap(~subset, ncol=1, labeller = labeller(subset = subset_labels))
```

```{r results-time-4, dev="png", dpi=300, fig.height=6, fig.cap="\\label{fig:results-stable-fixation}\\textbf{Stable polyploid fixation:} changes in relative diploid-polyploid abundances that culminate in stable polyploid fixation. Each point represents the 1 of 1000 simulations. Points are coloured from black to green based on the rate of nonreduction; simulations that ran with higher rates are brighter green. The dotted black line highlights a frequency of half the population capacity (0.5).", warning=FALSE}
subset_labels <- c("diploids", "sterile polyploids", "viable polyploids")
names(subset_labels) <- c("diploid_adults", "sterile_polyploid_adults", "viable_polyploid_adults")
            
simulations %>%
  filter(ID %in% stable_polyploid_winners$ID) %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::mutate(
    viable_polyploid_adults = polyploid_adults - sterile_polyploid_adults
  ) %>%
  tidyr::gather(subset, count, c(diploid_adults, sterile_polyploid_adults, viable_polyploid_adults)) %>%
  dplyr::filter(count != 0) %>%
  ggplot(aes(x = gen, y = (count*100000)/(adults*100000), colour = ploidy_rate)) +
  geom_hline(yintercept = .5, linetype = "dotted") +
  geom_point(alpha = 1, size = .5) +
  scale_colour_gradient(
    low = "black", high = ploidy_colours[1], 
    name = "Rate of\nnonreduction",
    limits = c(0, 0.5)
  ) +
  theme_classic() + theme_ploidy +
  scale_x_continuous(limits = c(0, 150)) +
  scale_y_continuous(breaks = seq(0, 1, 0.5)) +
  xlab("Time (generations)") +
  ylab("Relative abundance") +
  facet_wrap(~subset, ncol=1, labeller = labeller(subset = subset_labels))
```

## Sterile and viable polyploids

At low rates of nonreduction sterile polyploids were expected to become established at a relative abundance which roughly equalled the rate of nonreduction. In fact, their abundance was consistently double that. Figure \ref{fig:results-polyploids} shows how the established relative abundances of diploids and polyploids (taken from populations which reached the end of the simulations) related to the rate of nonreduction. As the rate of nonreduction was increased up to 0.5, the predictability of the relative abundance of sterile polyploids was expected to vary, either exceeding the rate of nonreduction or decreasing below it. As rate was increased, the relative abundace of established sterile polyploids began to climb steeply until the rate reached the threshold of 0.152. After this point, sterile polyploids no longer became established in any simulations which did not result in extinction. However, figures \ref{fig:results-extinction} and \ref{fig:results-unstable-fixation} show that there were points during population instability where sterile polyploids frequently accounted for around three quarters of the population.

A relative abundance of viable polyploids exceeding half the rate of nonreduction was expected to signal a dissapearance of sterile polyploids and allow fixation of polyploidy to persist. Figure \ref{fig:results-polyploids} shows that this was true. Under a rate of 0.152 very low relative abundances of viable polyploids become established in the system. As the rate of nonreduction increased, however, the shape of this relationship began to look exponential. Viable polyploids became established at relative abundances increasingly closer to half the rate of nonredcution. As the shape of this relationship reached a steep curve and looked set to hit the mark of half the rate, instability ensued: coexistence was no longer achieved (sterile polyploids and diploids no longer became established). Instead, there was only extinction or polyploid fixation.

```{r, results-polyploids, dev="png", dpi=300, fig.height=4, fig.cap="\\label{fig:results-polyploids}\\textbf{Relative abundance in repsonse to the rate of nonreduction:} Points represent relative abundance of sterile polyploids (red) or viable polyploids (blue) persisting in the population at generation 1000. Each point represents one of the 545 simulations which avoided total extinction out of 1000 instances. The solid black line shows where relative abundance equals the rate of nonreduction, the black dashed line below marks an abundance equal to half the rate, and the black dashed line above marks one of double. The dotted red line marks the threshold rate of nonreduction (0.152) where the probable outcome of simulations changed. The dotted blue line marks the rate above which the fixated viable polyploids were established in a stable manner around the carrying capacity (0.464)."}
simulations[which(simulations$ID %in% success), ] %>%
  tidyr::unnest(cols = c(data)) %>%
  dplyr::filter(gen == 1000) %>%
  dplyr::mutate(viable_polyploid_adults = polyploid_adults - sterile_polyploid_adults) %>%
  dplyr::select(-total) %>%
  tidyr::gather(subset, count, c(diploid_adults, sterile_polyploid_adults:viable_polyploid_adults)) %>%
  dplyr::filter(count != 0) %>%
  ggplot(aes(x = ploidy_rate, y = (count*100000)/(adults*100000), colour = subset)) +
  geom_point(alpha = .5, size = 1) +
  geom_vline(xintercept = 0.152, colour = ploidy_colours[4], linetype = "dotted") +
  geom_vline(xintercept = 0.464, colour = ploidy_colours[3], linetype = "dotted") +
  geom_line(aes(y = ploidy_rate), data = nonreduction, colour = "black") +
  geom_line(aes(y = ploidy_rate/2), data = nonreduction, colour = "black", linetype = "dashed") +
  geom_line(aes(y = ploidy_rate*2), data = nonreduction, colour = "black", linetype = "dashed") +
  scale_colour_manual(
    values = ploidy_colours[c(5,4,2)], 
    labels = c("diploid", "sterile polyploid", "viable polyploid"),
    name = ""
  ) +
  xlab("Rate of nonreduction") +
  ylab("Relative abundance") +
  theme_classic() + theme_ploidy
```


## Fixation of polyploidy

Fixation of polyploidy took one of two routes; 1) stable, or; 2) unstable. The pattern of demography change over time for these simulations was very similar (see figures \ref{fig:results-unstable-fixation} and \ref{fig:results-stable-fixation}). There was a period of polyploid growth where the relative abundance of viable polyploids exceeded that of the sterile polyploids, alongside a decline in diploids. Then a period where sterile and viable polyploids compete as viable polyploids increase in abundance until they account for 100% of the population. Fixation never occurred below a nonreduction rate of 0.152 (figure \ref{fig:results-polyploids}). Over that rate there was a 31.4% chance of polyploid fixation, and a 0.6% percent chance that this would be stable. At a rate of 0.464, the chance of achieving stable fixation increased to 7.0%, but the chance of unstable fixation dropped to 21.1%.

## Response of pioneering versus established diploid populations 

None of the outcomes (coexistence, extinction, or fixation) were more probable for either pioneering or established starting populations. Figure \ref{fig:results-hist-2} shows that roughly even numbers of simulations fall on either side of this scale for all outcomes. A minor density-dependent effect was seen in the speed with which simulations reached their fate; small starting populations took longer to stabilise or went extinct more quickly. Figures \ref{fig:results-outcomes-N} and \ref{fig:results-outcomes-rate} incorporate a time component; generation at which the outcome (coexistence, extinction, or fixation) was reached. Together, they show that the speed of achieving coexistence was influenced more by starting population size than by nonreduction rate; pioneering populations took longer to stabilise as the populations were further away from K. The converse was true for unstable fixation which was influenced by nonreduction rate rather than starting population size; the speed of reaching fixation above rates of 0.152 increased with rate no matter whether starting populations were pioneering or established. Total extinction was affected by both; rate influenced the speed of extinction most, but established populations buffered against this somewhat and delayed the effect. 

```{r results-hist-2, dev="png", dpi=300, fig.height=3, fig.cap="\\label{fig:results-hist-2}textbd{Chance of outcome depending on starting population size:} Each bin shows the number of simulations (from 1000 instances) within that group of starting population frequencies that had a particular outcome. Starting population frequency is the number of adults at the beginning of the simulaiton relative to carrying capacity (K), making the scale pioneering (close to zero) to established (close to 1). The dotted red line marks the halfway point between pioneering and established (0.5). Simulations ending with stable coexistence (top left) had a total adult population around the carrying capacity (100000) with both diploids and polyploids surviving throughout all 1000 generations. Simulations with total extinction (top right) had neither diploid or adults surviving to 1000 generations. Those where the relative abundance of polyploids persisting through to 1000 generations was 1, were considered fixated. Simulations ending in unstable polyploid fixation (bottom left) had polyploid populations of only 1 or 2 individuals (0.00001-0.00002% of the carrying capacity). Simulations ending with stable polyploid fixation (bottom right) had a total adult population around the carrying capacity (100000) which survived through to 1000 generations."}
outcomes %>%
  filter(outcome != "control") %>%
  ggplot(aes(x = starting_N/100000)) + 
    geom_histogram(colour = "white", bins = 50) +
    geom_vline(xintercept = 0.5, colour = ploidy_colours[4], linetype = "dotted") +
    scale_fill_manual(values = ploidy_colours[c(1,4,3,2)], name = "") +
    xlab("Rate of nonreduction") +
    ylab("Number of simulations") +
    theme_classic() + theme_ploidy +
    facet_wrap(~outcome, ncol = 2)
```



```{r results-outcomes-N, dev="png", dpi=300, fig.height=5.25, fig.cap="\\label{fig:results-outcomes-N}\\textbf{Speed of outcome depending on starting population size:} Starting population frequency is the number of adults at the beginning of the simulaiton relative to carrying capacity (K), making the scale pioneering (close to zero) to established (close to 1). Each point represents one of 1000 simulations, divided into panels by the outcome observed; stable coexistence (top left), total extinction (top right), unstable polyploid fixation (bottom left), and stable polyploid fixation (bottom right). Stability was defined as having a total population frequency that reached 1 and persisted around that point for the entire simulation. Total extinction meant that neither diploids or polyploid survived. Unstable polyploid fixation describes simulations where diploids went extinct but the remaining polyploids persisted at very low frequencies (around 0.00001). Points are coloured by the rate of nonreduction; low rates are black but as the rate increases the points become green."}
outcomes %>%
  filter(outcome != "control") %>%
  ggplot2::ggplot(aes(x = starting_N/100000, y = gen_reached, colour = ploidy_rate)) +
    ggplot2::geom_point(alpha = 1, size = 1) +
    ggplot2::theme_classic() + theme_ploidy +
    ggplot2::xlab("Starting population frequency") +
    ggplot2::ylab("Generation outcome reached") +
    ggplot2::scale_x_continuous(breaks = seq(0, 1, .5)) +
    scale_colour_gradient(
      low = "black", high = ploidy_colours[1], 
      name = "Rate of\nnonreduction"
    ) +
    facet_wrap(~outcome, ncol=2)
```

```{r results-outcomes-rate, dev="png", dpi=300, fig.height=5.25, echo=FALSE, message=FALSE, fig.cap="\\label{fig:results-outcomes-rate}\\textbf{Speed of outcome depending on rate of nonreduction:} each point represents one of 1000 simulations, divided into panels by the outcome observed; stable coexistence (top left), total extinction (top right), unstable polyploid fixation (bottom left), and stable polyploid fixation (bottom right). Stability was defined as having a total population frequency that reached 1 and persisted around that point for the entire simulation. Total extinction meant that neither diploids or polyploid survived. Unstable polyploid fixation describes simulations where diploids went extinct but the remaining polyploids persisted at very low frequencies (around 0.00001). Points are coloured by starting population frequency; well established starting populations are black but as the frequency of individuals within them reduces the points become red so that the smallest pioneering populations are the brightest. Starting population frequency is the number of adults at the beginning of the simulaiton relative to K, making the scale pioneering (close to zero) to established (close to 1). The dotted red line marks the threshold rate of nonreduction (0.152) which divided outcome probability."}
outcomes %>%
  filter(outcome != "control") %>%
  ggplot2::ggplot(aes(x = ploidy_rate, y = gen_reached, colour = starting_N/100000)) +
    geom_point(alpha = 1, size = 1.5) +
    geom_vline(xintercept = 0.152, colour = ploidy_colours[4], linetype = "dotted") +
    xlab("Rate of nonreduction") +
    ylab("Generation outcome reached") +
    scale_x_continuous(limits = c(0, 0.5)) +
    theme_classic() + theme_ploidy +
    scale_colour_gradient(
      low = ploidy_colours[4], high = "black",
      name = "Starting population\nfrequency"
    ) +
    facet_wrap(~outcome, ncol = 2)
    
```
